---
title: "Statistical Analysis of the N170"
author: "Ling"
date: "1/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(tidyverse)
library(knitr) # kable
library(gplots) # plot means with error bar

```

## Executive Summary   

Dataset is from this term's EEG course, 11 participants' N170 data included. Data was recorged from classic face recognition experiment, which has 2 variables and 4 conditions: ` face gender (female vs. male)` and `face orientation (upright vs. inverted)`.     

Identify the peak as the most negative point at both sides, and measure the *latency* of this peak. The *amplitude* was measured as the average voltage within the 120-220 time window.     

Brain region was used as a constrain to measure the N170 over the areas where it is expected to occur:  
  * Left: 58 + 59 + 65 + 69  
  * Right: 89 + 90 + 91 + 96    

So, firstly, an average across these channels to make sure that individual variability is taken into account. 

Secondly and mainly, an ANOVA would be run separately on the *latency* and the *amplitude* of the N170 using: 
  * hemisphere (left, right),  
  * gender (male, female), and  
  * face orientation (upright, inverted)   
      as within-subject factors.    

```{r data import and tidy, include=FALSE}
# data import 
N170_ampli <- read_delim("~/Desktop/Ling_R_final/N170_amplitude.csv", delim = ";")
N170_latency <- read_delim("~/Desktop/Ling_R_final/N170_latency.csv", delim = ";")

# tidy amplitude dataset 
tidy_N170_ampli <- N170_ampli %>%
  # average on left & right hemisphere electrodes
  transmute(female_inv_L = (Female_inv_L58 + Female_inv_L59 + Female_inv_L65 + Female_inv_L69)/4,
         female_inv_R = (Female_inv_R89 + Female_inv_R90 + Female_inv_R91 + Female_inv_R96)/4,
         female_up_L = (Female_up_L58 + Female_up_L59 + Female_up_L65 + Female_up_L69)/4,
         female_up_R = (Female_up_R89 + Female_up_R90 + Female_up_R91 + Female_up_R96)/4,
         male_inv_L = (Male_inv_L58 + Male_inv_L59 + Male_inv_L65 + Male_inv_L69)/4,
         male_inv_R = (Male_inv_R89 + Male_inv_R90 + Male_inv_R91 + Male_inv_R96)/4,
         male_up_L = (Male_up_L58 + Male_up_L59 + Male_up_L65 + Male_up_L69)/4,
         male_up_R = (Male_up_R89 + Male_up_R90 + Male_up_R91 + Male_up_R96)/4) %>%   
  gather(key = gender_faceOri_hemi, value = amplitude, female_inv_L:male_up_R, factor_key = TRUE) %>% 
  separate(col = gender_faceOri_hemi, into = c("gender", "face_ori", "hemisphere"))

# tidy latency dataset 
latency <- N170_latency %>%
  transmute(female_inv_L = (Female_inv_L58 + Female_inv_L59 + Female_inv_L65 + Female_inv_L69)/4,
            female_inv_R = (Female_inv_R89 + Female_inv_R90 + Female_inv_R91 + Female_inv_R96)/4,
            female_up_L = (Female_up_L58 + Female_up_L59 + Female_up_L65 + Female_up_L69)/4,
            female_up_R = (Female_up_R89 + Female_up_R90 + Female_up_R91 + Female_up_R96)/4,
            male_inv_L = (Male_inv_L58 + Male_inv_L59 + Male_inv_L65 + Male_inv_L69)/4,
            male_inv_R = (Male_inv_R89 + Male_inv_R90 + Male_inv_R91 + Male_inv_R96)/4,
            male_up_L = (Male_up_L58 + Male_up_L59 + Male_up_L65 + Male_up_L69)/4,
            male_up_R = (Male_up_R89 + Male_up_R90 + Male_up_R91 + Male_up_R96)/4) %>%
  gather(key = gender_faceOri_hemi, value = latency, female_inv_L:male_up_R, factor_key = TRUE) %>% 
  separate(col = gender_faceOri_hemi, into = c("gender", "face_ori", "hemisphere")) %>% 
  select(latency)

# add latency column 
N170_total <- cbind(latency, tidy_N170_ampli) 

```

## Descritives Statistics   

Extracting contidition means for amplitude and latency, respectively.  

```{r descritives stat, echo=FALSE}
# descritives stat for amplitude 
desc_ampli <- N170_ampli %>%
  # average on left & right hemisphere electrodes
  transmute(female_inv_L = (Female_inv_L58 + Female_inv_L59 + Female_inv_L65 + Female_inv_L69)/4,
         female_inv_R = (Female_inv_R89 + Female_inv_R90 + Female_inv_R91 + Female_inv_R96)/4,
         female_up_L = (Female_up_L58 + Female_up_L59 + Female_up_L65 + Female_up_L69)/4,
         female_up_R = (Female_up_R89 + Female_up_R90 + Female_up_R91 + Female_up_R96)/4,
         male_inv_L = (Male_inv_L58 + Male_inv_L59 + Male_inv_L65 + Male_inv_L69)/4,
         male_inv_R = (Male_inv_R89 + Male_inv_R90 + Male_inv_R91 + Male_inv_R96)/4,
         male_up_L = (Male_up_L58 + Male_up_L59 + Male_up_L65 + Male_up_L69)/4,
         male_up_R = (Male_up_R89 + Male_up_R90 + Male_up_R91 + Male_up_R96)/4) %>%   
  gather(key = gender_faceOri_hemi, value = amplitude, female_inv_L:male_up_R, factor_key = TRUE) %>% 
  group_by(gender_faceOri_hemi) %>% 
  summarise(N=length(amplitude),
            Mean_amplitude=mean(amplitude, na.rm=TRUE), 
            SD=sd(amplitude, na.rm=TRUE), 
            Min=min(amplitude, na.rm=TRUE), 
            Max=max(amplitude, na.rm=TRUE), 
            SEM = sd(amplitude)/sqrt(length(amplitude)),
            '% Missing'=100*length(which(is.na(amplitude)))/n())

kable(desc_ampli, digits=2, format="pandoc", caption="Table 1: Descriptive Statistics for N170 Amplitude")

# descritives stat for latency
desc_latency <- N170_latency %>%
  transmute(female_inv_L = (Female_inv_L58 + Female_inv_L59 + Female_inv_L65 + Female_inv_L69)/4,
            female_inv_R = (Female_inv_R89 + Female_inv_R90 + Female_inv_R91 + Female_inv_R96)/4,
            female_up_L = (Female_up_L58 + Female_up_L59 + Female_up_L65 + Female_up_L69)/4,
            female_up_R = (Female_up_R89 + Female_up_R90 + Female_up_R91 + Female_up_R96)/4,
            male_inv_L = (Male_inv_L58 + Male_inv_L59 + Male_inv_L65 + Male_inv_L69)/4,
            male_inv_R = (Male_inv_R89 + Male_inv_R90 + Male_inv_R91 + Male_inv_R96)/4,
            male_up_L = (Male_up_L58 + Male_up_L59 + Male_up_L65 + Male_up_L69)/4,
            male_up_R = (Male_up_R89 + Male_up_R90 + Male_up_R91 + Male_up_R96)/4) %>%
  gather(key = gender_faceOri_hemi, value = latency, female_inv_L:male_up_R, factor_key = TRUE) %>% 
  group_by(gender_faceOri_hemi) %>% 
  summarise(N=length(latency),
            Mean_latency=mean(latency, na.rm=TRUE), 
            SD=sd(latency, na.rm=TRUE), 
            Min=min(latency, na.rm=TRUE), 
            Max=max(latency, na.rm=TRUE), 
            SEM = sd(latency)/sqrt(length(latency)),
            '% Missing'=100*length(which(is.na(latency)))/n())

kable(desc_latency, digits=2, format="pandoc", caption="Table 2: Descriptive Statistics for N170 Latency")
```

## EDA Plots  

### Amplitude and latency interaction plot   
```{r plot1, echo=FALSE}
# wonder the relation between amplitude and latency, just for fun
N170_total %>%
  ggplot() +
  aes(x =latency, y = amplitude, color = gender, size = hemisphere, label = face_ori) +
  geom_point(alpha = 0.4) +
  theme_bw() +
  ggtitle("Interaction between Amplitude and Latency")
```

### Amplitude plot    
```{r plot amplitude, echo=FALSE}
# ok, seriously plot
desc_ampli %>% 
  separate(col = gender_faceOri_hemi, into = c("gender", "face_orientation", "hemisphere")) %>% 
  ggplot() +
  aes(x = face_orientation, y = Mean_amplitude, fill = face_orientation) +
  geom_bar(position = position_dodge(), stat = "identity", width = .5) +
  geom_errorbar(aes(ymin = Mean_amplitude - SEM, ymax = Mean_amplitude + SEM), width = .1, position = position_dodge(.9)) +
  facet_grid(~gender + hemisphere) +
  ylab("Voltage(µV)") +
  theme_bw() + 
  ggtitle(paste("Amplitude for various face orientation\n(bars indicating SEM, N =", desc_ampli$N, ")"))
```

### Latency plot       
```{r plot latency, echo=FALSE}
desc_latency %>% 
  separate(col = gender_faceOri_hemi, into = c("gender", "face_orientation", "hemisphere")) %>% 
  ggplot() +
  aes(x = face_orientation, y = Mean_latency, fill = face_orientation) +
  geom_bar(position = position_dodge(), stat = "identity", width = .5) +
  geom_errorbar(aes(ymin = Mean_latency - SEM, ymax = Mean_latency + SEM), width = .1, position = position_dodge(.9)) +
  facet_grid(~gender + hemisphere) +
  ylab("Latency (ms)") +
  theme_bw() + 
  ggtitle(paste("Latency for various face orientation\n(bars indicating SEM, N =", desc_latency$N, ")"))
```

## Inferential Analysis  

### ANOVA on Amplitude   

1. **Two-way interactions plots**  
    A visual way to see hether different combinations of the levels of the factors differently affect the average amplitude.  
    
```{r ampli two way interaction plot, echo=FALSE}
# plot two way interaction
am <- par(mfrow = c(3, 1))
with(N170_total, {
  interaction.plot(gender, face_ori, amplitude)
  interaction.plot(gender, hemisphere, amplitude)
  interaction.plot(face_ori, hemisphere, amplitude)
}
)
par(am)
```
  
There may be an interaction between `face orientaion` and `hemisphere`, meaning that `face orientaion` might affect the `amplitude` differently based on whether it is combined with left or right hemisphere.   

2. **Fit a Model**  
   Build a full model to see whether the differences between means plotted are significant or not.  
```{r ampli full model, echo=FALSE}
# amplitude ANOVA
ampli_model <- lm(amplitude ~ gender * face_ori * hemisphere, data = N170_total)
ampli_anova <- anova(ampli_model)
kable(ampli_anova, digits=2, format="pandoc", caption="Table 3: ANOVA on N170 Amplitude")
```
Only main effect of `face orientation` is significant, F(1, 80) = 11.62, p < .001.  

3. **Diagnostic plots**  
  
```{r ampli diagnostic plot, echo=FALSE}
#1 get full model again
m1 <- aov(amplitude ~ gender * face_ori * hemisphere, data = N170_total)
# Residual analysis of 3-way Anova
op <-  par(mfrow = c(2, 2))
plot(m1)
par(op)
```
   
4. **Model Update**  
   4.1 Firstly, remove 3-way interaction  
   4.2 Then, remove 2-way interaction  
   4.3 Check if the three removed effects together are still not significant  
   4.4 Obtain a final model with significant effect only  
```{r ampli model update, echo=FALSE}
## model update
#2 remove 3-way interaction, use updata() 
m2 <- update(m1, . ~ . - gender:face_ori:hemisphere)
summary(m2)

#3 remove 2-way interaction, always use updata()
m3 <- update(m2, .~gender+face_ori+hemisphere)
summary(m3)

#4 check if the three removed effects together are still not significant using anova()
anova(m2, m3) # not significant

#5 obtain the final model with significant effects only
final_m <- aov(amplitude ~ face_ori, data = N170_total)
summary(final_m)

```
5. **Visualise the results**   
```{r ampli line plot result, echo=FALSE}
# visualizing results
attach(N170_total)

# plot means with error bars
plotmeans(amplitude ~ face_ori, xlab="Face Orientation",
  ylab="Amplitude", main="Mean Plot\nwith 95% CI")
```
  
As in the mean plot, upright face elicited larger N170 amplitude than inverted face. 
  
### ANOVA on Latency   

1. **Two-way interactions plots**  
```{r late two way interaction plot, echo=FALSE}
# plot two way interaction
am <- par(mfrow = c(3, 1))
with(N170_total, {
  interaction.plot(gender, face_ori, latency)
  interaction.plot(gender, hemisphere, latency)
  interaction.plot(face_ori, hemisphere, latency)
}
)
par(am)
```
  
There may be two interactions between `face orientaion` vs. `gender`, and `face orientaion` vs. `hemisphere`.   

2. **Fit a Model**  
   Build a full model to see whether the differences between means plotted are significant or not.  
```{r late full model, echo=FALSE}
# latency ANOVA
late_model <- lm(latency ~ gender * face_ori * hemisphere, data = N170_total)
late_anova <- anova(ampli_model)
kable(late_anova, digits=2, format="pandoc", caption="Table 4: ANOVA on N170 Latency")
```
  
No significant results were found.    


## Conclusion  

Mainly, a 3-way within-subject ANOVA was run separately on the *amplitude* and *latency* of the N170. Only the main effect of `face orientation` on *amplitude* was found,  F(1, 80) = 11.62, p < .001, suggesting `face orientation` has an impact on *amplitude*. Furthermore, N170 shows sensitivity to human face inversion, which is in line with previous findings.




